# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

The code below will check if the data file exists in the current folder. If not, the script will download the data from internet. Thus, please make sure you are online if you don't have the data in place, or the code might fail downloading the required data.

This part of code is tested under Linux. You might need to modify the *method* parameter for the download.file() function to suit your system.

```{r}

dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

### Download data
if (file.exists("./activity.zip") || file.exists("./activity.csv")) {
  message("== Data file (zip or csv format) exists, skipping downloading step \r")
  
} else {

  message("== Downloading data file...\r")
  download.file(url=dataURL,destfile="./activity.zip",method="wget")
  
}

### Extract data
if (file.exists("./activity.csv")) {
  message("== Data file (csv format) exists, skipping extraction...\r")
} else {
  message("== Extracting data ... \r")
  unzip(zipfile="./activity.zip",exdir="./",)  
}


### Load data
data <- read.csv("./activity.csv")

data$date <- as.Date(data$date,format="%Y-%m-%d")
data$interval <- as.factor(data$interval)

```

## What is mean total number of steps taken per day?

```{r}

library("plyr")
dfPerDate <- ddply(data,~date,summarise,sumSteps=sum(steps,na.rm=TRUE),meanSteps=mean(steps,na.rm=TRUE),sdSteps=sd(steps,na.rm=TRUE))

meanTotalStepsPerDay <- mean(dfPerDate$sumSteps,na.rm=TRUE)
medianTotalStepsPerDay <- median(dfPerDate$sumSteps,na.rm=TRUE)

hist(dfPerDate$sumSteps,col="red", xlab="",main="Total steps per day",breaks=10)

```

**Fig. 1:** Histogram of the total number of steps taken each day

The **mean** and **median** of total number of steps taken per day are **`r meanTotalStepsPerDay`** and **`r medianTotalStepsPerDay`**, respectively.


## What is the average daily activity pattern?

```{r}
dfPerInterval <- ddply(data,~interval,summarise,sumSteps=sum(steps,na.rm=TRUE),meanSteps=mean(steps,na.rm=TRUE),sdSteps=sd(steps,na.rm=TRUE))

intervalWithMostSteps <- dfPerInterval$interval[dfPerInterval$meanSteps == max(dfPerInterval$meanSteps)]

with(dfPerInterval, plot(interval,meanSteps, type = "l",xlab="Interval",ylab="Mean steps",main="Mean steps per 5-min interval"))
with(dfPerInterval, lines(interval,meanSteps))

```
**Fig. 2:** time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.

The `r intervalWithMostSteps` interval contains the maximum number of steps.


## Imputing missing values

```{r}

NumMissing <- sum(is.na(data$steps))

# imputing data
imputtedData <- data

# Replacing the missing values with the mean for that 5-minute interval across days
for (i in which(is.na(data$steps))) {
  imputtedData$steps[i] <- dfPerInterval$meanSteps[which(dfPerInterval$interval==imputtedData$interval[i])]
}

newdfPerDate <- ddply(imputtedData,~date,summarise,sumSteps=sum(steps,na.rm=TRUE),meanSteps=mean(steps,na.rm=TRUE),sdSteps=sd(steps,na.rm=TRUE))

newmeanTotalStepsPerDay <- mean(newdfPerDate$sumSteps,na.rm=TRUE)
newmedianTotalStepsPerDay <- median(newdfPerDate$sumSteps,na.rm=TRUE)

hist(newdfPerDate$sumSteps,col="red", xlab="",main="Total steps per day -- Imputted data",breaks=10)

```

**Fig. 3:** Histogram of the total number of steps taken daily for the imputted data.

There are `r NumMissing` missing values in the dataset. The missing values were then replaced with the mean for that 5-minute interval across days.


The **mean** and **median** of total number of steps taken per day are **`r newmeanTotalStepsPerDay`** and **`r newmedianTotalStepsPerDay`**, respectively.

The mean and median of the total number of steps taken daily increased after data imputing.

## Are there differences in activity patterns between weekdays and weekends?
